---
title: "ChangeUp"
author: "Taylor Villahermosa"
output: pdf_document
---
\setlength{\belowdisplayskip}{0pt} \setlength{\belowdisplayshortskip}{0pt}
\setlength{\abovedisplayskip}{0pt} \setlength{\abovedisplayshortskip}{0pt}
```{r echo=FALSE, message=FALSE}
library(dplyr)
library(DBI)
library(RSQLite)
library(sqldf)
library(ggplot2)
```
```{r creating ultimate}
ultimate <- read.csv("/Users/taylorvillahermosa/Desktop/NewUltimateBBScore.csv")

# Add strike and whiff count to ultimate
ultimate <- ultimate %>% 
  mutate(Strike = ifelse(PitchCall %in% c("StrikeSwinging","StrikeCalled",
                                          "FoulBall","InPlay"), 1, 0)) %>%
  mutate(Whiff = ifelse(PitchCall=="StrikeSwinging",1,0))

# Create data base
master_db <- dbConnect(RSQLite::SQLite(), "master_db.sqlite")
dbWriteTable(master_db,"ULTIMATE",ultimate,overwrite=TRUE)

master <- as.data.frame(dbGetQuery(master_db, 'SELECT PitchNo, Batter, Pitcher, 
                                                  TaggedPitchType, RelSpeed, 
                                                  PitchCall, Strikes, InducedVertBreak,
                                                  Strike, Whiff, ExitSpeed, BBScore
                                                  FROM ULTIMATE'))
# Remove undefined pitches
x <- subset(master, !(is.na(Batter)) | is.na(TaggedPitchType))

# Create new columns for Diff and Prop
x$VBDiff <- NA
x$VBProp <- NA
x$VeloDiff <- NA
x$VeloProp <- NA

for(i in 1:(nrow(x)-1)){  
  if (x[i,"TaggedPitchType"] == "Fastball" &&    # 1st pitch FB
      x[i+1,"TaggedPitchType"] == "ChangeUp" &&  # 2nd pitch U
      x[i,"Batter"] == x[i+1,"Batter"] &&        # same at bat
      x[i,"Pitcher"] == x[i+1,"Pitcher"])
        {
        x[i+1,"VBDiff"] <- x[i,"InducedVertBreak"] - x[i+1,"InducedVertBreak"]
        x[i+1,"VBProp"] <- 1 - (x[i+1,"InducedVertBreak"]/x[i,"InducedVertBreak"])
        x[i+1,"VeloDiff"] <- x[i,"RelSpeed"] - x[i+1,"RelSpeed"] 
        x[i+1,"VeloProp"] <- 1 - (x[i+1,"RelSpeed"]/x[i,"RelSpeed"])
        }
}

# Remove non-ChangeUps and outliers
hi <- x


x <- subset(x, !is.na(VeloDiff))# & VBProp > 0) 
x <- subset(x, VBProp > 0)
x_rounded <- x %>%
  mutate(VeloDiff = round((VeloDiff/2))*2) %>%  # velo diff to 2
  mutate(VeloProp = (round(VeloProp/2, digits = 2)*2)) %>%  # velo prop to 0.02 (2%)
  mutate(Velo = round((RelSpeed)/5, digits = 1)*5) %>%  # velo to 0.5
  mutate(VBDiff = round((VBDiff))) %>%  # vb diff to 1
  mutate(VBProp = (round(VBProp/10, digits=2)*10))  # vb prop to 0.1 (10%)

```
\newpage  
## Subsetting
```{r subset data}
dbWriteTable(master_db,"subset",x,overwrite=TRUE)
dbWriteTable(master_db,"subset2",x_rounded,overwrite=TRUE)

# VB Prop and VB Diff
t <- as.data.frame(dbGetQuery(master_db, 'SELECT VeloDiff, VBDiff,
                                            round(100*VeloProp,1) AS VeloProp,
                                            round(100*VBProp,1) AS VBProp,
                                            round((ExitSpeed),1) AS "ExitVelo"
                                            FROM subset'))

# VB Diff (avg ev & whiff %)
dw <- as.data.frame(dbGetQuery(master_db, 'SELECT VBDiff,
                                            COUNT(*) AS "PitchCount", 
                                            round(SUM(Whiff)/COUNT(*)*100,1) AS "Whiff%",
                                            round(AVG(ExitSpeed),1) AS "ExitVelo"
                                            FROM subset2 GROUP BY VBDiff'))
dw <- subset(dw, PitchCount > 100)

# VB Prop (avg ev & whiff %)
pw <- as.data.frame(dbGetQuery(master_db, 'SELECT VBProp,
                                            COUNT(*) AS "PitchCount", 
                                            round(SUM(Whiff)/COUNT(*)*100,1) AS "Whiff%",
                                            round(AVG(ExitSpeed),1) AS "ExitVelo"
                                            FROM subset2 GROUP BY VBProp'))
pw <- subset(pw, PitchCount > 100)

# Velo Diff (avg ev & whiff %)
z <- as.data.frame(dbGetQuery(master_db, 'SELECT VeloDiff, 
                                            COUNT(*) AS "PitchCount", 
                                            round(SUM(Whiff)/COUNT(*)*100,1) AS "Whiff%",
                                            round(AVG(ExitSpeed),1) AS "ExitVelo"
                                            FROM subset2 GROUP BY VeloDiff'))
z <- subset(z, PitchCount > 100)

# Velo Prop (avg ev & whiff %)
z2 <- as.data.frame(dbGetQuery(master_db, 'SELECT VeloProp,
                                            COUNT(*) AS "PitchCount", 
                                            round(SUM(Whiff)/COUNT(*)*100,1) AS "Whiff%",
                                            round(AVG(ExitSpeed),1) AS "ExitVelo"
                                            FROM subset2 GROUP BY VeloProp'))
z2 <- subset(z2, PitchCount > 100)

# Plot Labels
D <- "Velo Differential"
P <- "Velo Proportion"
VD <- "VB Differential"
VP <- "VB Proportion"
W <- "Whiff %"
E <- "Exit Velo"
```
```{r}
# ev velo prop
grape <- x_rounded %>% filter(VeloProp >= 0.04 & VeloProp < .145)
turtle <- grape %>% group_by(VeloProp) %>% sample_n(100) 
dolphin <- turtle %>% summarise(ExitVelo = mean(na.omit(ExitSpeed)))


ggplot(dolphin,aes(VeloProp,ExitVelo))+
  geom_smooth(method = "loess",color="red",fill="lightskyblue",size=0.2)+
  geom_point(size=0.2)+
  ggtitle(paste0("Avg ",E," vs. ",P))+labs(x=P,y=E)+
  theme(plot.title = element_text(hjust =0.5))

fit <- lm(ExitVelo~VeloProp, data = dolphin)
summary(fit)

a1 <- x_rounded %>% filter(VeloDiff >= 3.5 & VeloDiff < 13)
a2 <- a1 %>% group_by(VeloDiff) %>% sample_n(100) 
a3 <- a2 %>% summarise(ExitVelo = mean(na.omit(ExitSpeed)))

ggplot(a3,aes(VeloDiff,ExitVelo))+
  geom_smooth(method = "glm",color="red",fill="lightskyblue",size=0.2)+
  geom_point(size=0.2)+
  ggtitle(paste0("Avg ",E," vs. ",D))+labs(x=D,y=E)+
  theme(plot.title = element_text(hjust =0.5))

#fit <- lm(ExitVelo~VeloDiff, data = a3)
#summary(fit)
```
\newpage
## FB-U Velocity Proportion/Differential
```{r fig.height = 8, fig.width=7}
sloth <- subset(na.omit(t), -100 < VeloProp & VeloProp < 100 & VeloDiff > 0)

# (1) ev vs velo diff
ev_velodiff  <- ggplot(sloth,aes(VeloDiff,ExitVelo))+
  geom_point(size=0.2)+
  ggtitle(paste0(E," vs. ",D))+
  theme(plot.title = element_text(hjust =0.5))

# (2) ev vs velo prop
ev_veloprop <- ggplot(sloth,aes(VeloProp,ExitVelo))+
  geom_point(size=0.2)+
  ggtitle(paste0(E," vs. ",P))+
  theme(plot.title = element_text(hjust =0.5))

# (3) avg ev vs velo diff
avg_ev_velodiff <- ggplot(z,aes(VeloDiff,ExitVelo))+
  geom_smooth(method = "loess",color="red",fill="lightskyblue",size=0.2)+
  geom_point(size=0.2)+
  ggtitle(paste0("Avg ",E," vs. ",D))+labs(x=D,y=E)+
  theme(plot.title = element_text(hjust =0.5))

# (4) avg ev vs velo prop
avg_ev_veloprop <- ggplot(z2,aes(VeloProp,ExitVelo))+
  geom_smooth(method = "loess",color="red",fill="lightskyblue",size=0.2)+
  geom_point(size=0.2)+
  ggtitle(paste0("Avg ",E," vs. ",P))+labs(x=P,y=E)+
  theme(plot.title = element_text(hjust =0.5))

# (5) whiff vs velo diff
w_velodiff <- ggplot(z,aes(VeloDiff,`Whiff%`))+
  geom_smooth(method = "glm",color="red",fill="lightskyblue",size=0.2)+
  geom_point(size=0.2)+
  ggtitle(paste0(W," vs. ",D))+labs(x=D,y=W)+
  theme(plot.title = element_text(hjust =0.5))

# (6) whiff vs velo prop
w_veloprop <- ggplot(z2,aes(VeloProp,`Whiff%`))+
  geom_smooth(method = "glm",color="red",fill="lightskyblue",size=0.2)+
  geom_point(size=0.2)+
  ggtitle(paste0(W," vs. ",P))+labs(x=P,y=W)+
  theme(plot.title = element_text(hjust =0.5))

cowplot::plot_grid(ev_velodiff,ev_veloprop,
                   avg_ev_velodiff, avg_ev_veloprop,
                   w_velodiff, w_veloprop,
                   ncol=2,labels=1:6)
```
\newpage
```{r}
# (5) whiff vs velo diff
fit5 <- lm(`Whiff%`~VeloDiff, data = z)
summary(fit5)
# (6) whiff vs velo prop
fit6 <- lm(`Whiff%`~VeloProp, data = z2)
summary(fit6)
```
\newpage
## Vertical Break 
```{r fig.height = 8, fig.width=7}
giraffe <- subset(na.omit(t), -100 < VBProp & VBProp < 100 & VBDiff > 0)

# (7) EV vs VBDiff 
ev_vbdiff <- ggplot(giraffe,aes(VBDiff,ExitVelo))+
  geom_point(size=0.2)+
  ggtitle(paste0(E," vs. ",VD))+
  theme(plot.title = element_text(hjust =0.5))

# (8) EV vs VBProp 
ev_vbprop <- ggplot(giraffe,aes(VBProp,ExitVelo))+
  geom_point(size=0.2)+
  ggtitle(paste0(E," vs. ",VP))+
  theme(plot.title = element_text(hjust =0.5))

# (9) avg ev vs vb diff
avg_ev_vbdiff <- ggplot(dw,aes(VBDiff,ExitVelo))+
  geom_smooth(method = "glm",color="red",fill="lightskyblue",size=0.2)+
  geom_point(size=0.2)+
  ggtitle(paste0("Avg ",E," vs. ",VD))+labs(x=VD,y=E)+
  theme(plot.title = element_text(hjust =0.5))

# (10) avg ev vs vb prop
avg_ev_vbprop <- ggplot(pw,aes(VBProp,ExitVelo))+
  geom_smooth(method = "glm",color="red",fill="lightskyblue",size=0.2)+
  geom_point(size=0.2)+
  ggtitle(paste0("Avg ",E," vs. ",VP))+labs(x=VP,y=E)+
  theme(plot.title = element_text(hjust =0.5))

# (11) Whiff vs VBDiff
w_vbdiff <- ggplot(dw,aes(VBDiff,`Whiff%`))+
  geom_smooth(method = "glm",color="red",fill="lightskyblue",size=0.2)+
  geom_point(size=0.2)+
  ggtitle(paste0(W," vs. ",VD))+ylab(W)+
  theme(plot.title = element_text(hjust =0.5))

# (12) Whiff vs VBProp
w_vbprop <- ggplot(pw,aes(VBProp,`Whiff%`))+
  geom_smooth(method = "glm",color="red",fill="lightskyblue",size=0.2)+
  geom_point(size=0.2)+
  ggtitle(paste0(W," vs. ",VP))+ylab(W)+
  theme(plot.title = element_text(hjust =0.5))

cowplot::plot_grid(ev_vbdiff, ev_vbprop, 
                   avg_ev_vbdiff, avg_ev_vbprop,
                   w_vbdiff, w_vbprop, ncol = 2,
                   labels = 7:12)
```
\newpage
```{r}
# (9) avg ev vs vb diff
fit9 <- lm(ExitVelo~VBDiff, data = dw)
summary(fit9)
# (10) avg ev vs vb prop
fit10 <- lm(ExitVelo~VBProp, data = pw)
summary(fit10)
```
\newpage
```{r}
# (11) whiff vs vb diff
fit11 <- lm(`Whiff%`~VBDiff, data = dw)
summary(fit11)
# (12) whiff vs vb prop
fit12 <- lm(`Whiff%`~VBProp, data = pw)
summary(fit12)
```  
\newpage


## Comparing 1st, 2nd, and 3rd ChangeUps with 2 strikes
```{r three change ups}
master <- read_excel("/Users/taylorvillahermosa/Desktop/good changeups.xlsx")
#master <- read.csv("/Users/taylorvillahermosa/Desktop/new good changeups.csv")

#y <- subset(master, PitchCall != "FoulBall")
y <- subset(y, HitType != "Bunt")

y$u1EV <- NA
y$u2EV <- NA
y$u3EV <- NA
y$u.num <- NA

# 1 ChangeUp
for(i in 1:(nrow(y)-2)){
  if (y[i,"TaggedPitchType"] != "ChangeUp" &&
      y[i+1,"TaggedPitchType"] == "ChangeUp" && 
      y[i+2,"TaggedPitchType"] != "ChangeUp")
      {
        y[i+1,"u1EV"] <- y[i+1,"ExitSpeed"]
        y[i+1,"u.num"] <- 1
      }
}

# 2 ChangeUps in a row
for(i in 1:(nrow(y)-3)){
  if (y[i,"TaggedPitchType"] != "ChangeUp" && 
      y[i+1,"TaggedPitchType"] == "ChangeUp" &&
      y[i+2,"TaggedPitchType"] == "ChangeUp" &&
      y[i+3,"TaggedPitchType"] != "ChangeUp"&&
      y[i,"Batter"] == y[i+1,"Batter"]) 
      {
        y[i+2,"u2EV"] <- y[i+2,"ExitSpeed"]
        y[i+1,"u.num"] <- 1
        y[i+2,"u.num"] <- 2
      }
}

# 3 ChangeUps in a row
for(i in 1:(nrow(y)-3)){
  if (y[i,"TaggedPitchType"] == "ChangeUp" && 
      y[i+1,"TaggedPitchType"] == "ChangeUp" &&
      y[i+2,"TaggedPitchType"] == "ChangeUp" &&
      y[i,"Batter"] == y[i+1,"Batter"] &&
      y[i+1,"Batter"] == y[i+2,"Batter"]) 
      {
        y[i+2,"u3EV"] <- y[i+2,"ExitSpeed"]
        y[i,"u.num"] <- 1
        y[i+1,"u.num"] <- 2
        y[i+2,"u.num"] <- 3
      }
}

c2 <- subset(y, Strikes == 2 & !is.na(u.num))

c2 <- c2 %>% 
  mutate(Whiff = ifelse(PitchCall=="StrikeSwinging",1,0))
```
\newpage
## Comparing Whiff Rates
```{r}

potato <- c2
master_db <- dbConnect(RSQLite::SQLite(), "master_db.sqlite")

dbWriteTable(master_db,"potato",potato,overwrite=TRUE)

turtle <- as.data.frame(dbGetQuery(master_db, 'SELECT "u.num",
                                            COUNT(*) AS "PitchCount", 
                                            SUM(Whiff) AS "WhiffCount",
                                            SUM(Whiff)/COUNT(*)*100 AS "Whiff%"
                                            FROM potato
                                            GROUP BY "u.num"'))

w1 <- subset(turtle$WhiffCount, turtle$u.num == 1)
w2 <- subset(turtle$WhiffCount, turtle$u.num == 2)
w3 <- subset(turtle$WhiffCount, turtle$u.num == 3)
n1 <- subset(turtle$PitchCount, turtle$u.num == 1)
n2 <- subset(turtle$PitchCount, turtle$u.num == 2)
n3 <- subset(turtle$PitchCount, turtle$u.num == 3)


c((whiff1 <- w1/n1),n1)  # 1st change up

c((whiff2 <- w2/n2),n2)  # 2nd change up

c((whiff3 <- w3/n3),n3)  # 3rd change up
```
### whiff rate 1 < whiff rate 2 < whiff rate 3
```{r}
# Whiff rate for 1st changeup is lower than 2nd with 2 strikes
w12 <- prop.test(x = c(w1, w2), n = c(n1, n2), alternative = "less")
c(w12$p.value,w12$conf.int)
```
$~$  
```{r}
# Whiff rate for 1st changeup is lower than 3rd with 2 strikes
w13 <- prop.test(x = c(w1, w3), n = c(n1, n3), alternative = "less")
c(w13$p.value,w13$conf.int)
```
$~$  
```{r}
# Whiff rate for 2nd changeup is lower than 3rd with 2 strikes
w23 <- prop.test(x = c(w2, w3), n = c(n2, n3), alternative = "less")
c(w23$p.value,w23$conf.int)
```  
\newpage
## Comparing Exit Velos
```{r}
c2 <- subset(c2, PitchCall != "FoulBall")

twoKev1 <- na.omit(c2$u1EV)
twoKev2 <- na.omit(c2$u2EV)
twoKev3 <- na.omit(c2$u3EV)

c(mean(twoKev1),length(twoKev1))  # 1st change up
c(mean(twoKev2),length(twoKev2))  # 2nd change up
c(mean(twoKev3),length(twoKev3))  # 3rd change up
```
### No significant differences between Exit Velos and # of changeups in a row
```{r}
t.test(twoKev1, twoKev2, alternative = "greater", var.equal = FALSE)$p.value
t.test(twoKev1, twoKev3, alternative = "two.sided", var.equal = FALSE)$p.value
t.test(twoKev2, twoKev3, alternative = "less", var.equal = FALSE)$p.value
```  